Choosing Color Palettes for Statistical Graphics

### 2007-10-31 pkg/paper/Palettes/CSDA.Rnw
possible addition to intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/colorspace@19 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\documentclass[shortnames]{Z}
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide}
\usepackage{amsfonts,rotating}

@@

Color is an integral element of graphical displays in general, and many statistical
graphics in particular. Statistical software packages typically provide various color
palettes and allow practitioners to employ these for visualizing data in various types
of displays. However, more often than not, there is relatively little guidance about
how to choose a suitable palette for a certain visualization task. Consequently,
sub-optimal color palettes are used quite frequently for color graphics displayed in journal articles
or on presentation slides which could easily be enhanced if the underlying information
were captured appropriately in the color schemes. Here, we address
this problem by suggesting a color selection strategy for visualizing both categorical
and numerical information by selecting colors along axes in a color space whose axes
can be matched with perceptual axes of the human visual system.

@@

Our first illustration is a heatmap, a very popular display for visualizing a scalar function of
two arguments. Here, we use a bivariate kernel density estimate \citep[for the Old
Faithful geyser eruptions data from][]{color:Azzalini+Bowman:1990}---other typical applications
include objective functions with two arguments \citep[as in][]{color:Gneiting+Sevcikova+Percival:2006}
or (physical) measurements on a 2-dimensional grid \citep[as in][]{color:Yang+Buckley+Dudoit:2002}.
Figure~\ref{fig:bkde} shows such a heatmap bringing out the relationship between the duration
of an eruption of the Old Faithful geyser in Yellowstone National Park 
and the waiting time for this eruption. It reveals a multi-modal distribution: short waiting
times (around 50 minutes) are typically followed by a long eruption (around 4 minutes)
whereas long waiting times (around 80 minutes) can be followed by either a long
or short eruption (around 4 minutes).

>>>>

\documentclass[shortnames]{Z}
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, echo=FALSE, results=hide}
\usepackage{amsfonts,rotating}
\usepackage{paralist}

@@

Color is an integral element of graphical displays in general, and many statistical
graphics in particular.  Furthermore, 
an increasing amount of research is being published and viewed online, making
it easier to include color in published plots.
However, selecting which colors to use for a statistical graphic is
not a trivial task.

There are three main obstacles to overcome:
\begin{compactitem}
\item
The colors should not be unappealing.  
\end{compactitem}

It is not necessary for the 
colors in a statistical plot to reflect the latest in fashion trends, 
but basic 
principles such as avoiding fully saturated colors should be
adhered to.  The requirement is not that the user should have
a degree in graphic design, but
that the software should allow the user to select colors and provide
control over basic properties such as how bright the colors are.

In order to be able to control basic perceptual features of colors,
it is necessary to work in a \emph{colorspace} that describes 
colors in terms of their perceptual properties.
This article will describe an appropriate colorspace for
selecting colors.

\begin{compactitem}
\item 
The colors in a statistical graphic
should cooperate with each other.  
\end{compactitem}

The typical purpose 
of color in a statistical graphic is to distinguish between different 
areas or symbols in the plot---to distinguish between different
groups or between different levels of a variable.  This means
that there will typically be several colors, or a \emph{palette}
of colors, used within a plot and 
that those colours should be related to each other.

Part of the solution to this problem is to work in an
appropriate colorspace, as in the previous point, so that 
motion within the colorspace and distances between colors 
have an intuitive and rational meaning.  However, 
there still remains the issue of providing basic navigation 
tools so that users can conveniently generate color
palettes.  The main part of this article will describe
algorithms for generating color palettes for common 
applications in statistical graphics.

\begin{compactitem}
\item
The colors should work everywhere.  
\end{compactitem}

In the ideal situation, colors 
should be selected so that they continue to work in any 
context.  For example, different areas of a plot should still be 
distiguishable when printed in black-and-white, or when the person
viewing the graphic is colour-blind.
This article will also briefly address these issues.

@@

Our first illustration is a heatmap, a very popular display for visualizing a scalar function of
two arguments. Here, we use a bivariate kernel density estimate \citep[for the Old
Faithful geyser eruptions data from][]{color:Azzalini+Bowman:1990}---other typical applications
include objective functions with two arguments \citep[as in][]{color:Gneiting+Sevcikova+Percival:2006}
or (physical) measurements on a 2-dimensional grid \citep[as in][]{color:Yang+Buckley+Dudoit:2002}.
Figure~\ref{fig:bkde} shows such a heatmap bringing out the relationship between the duration
of an eruption of the Old Faithful geyser in Yellowstone National Park 
and the waiting time for this eruption. It reveals a multi-modal distribution: short waiting
times (around 50 minutes) are typically followed by a long eruption (around 4 minutes)
whereas long waiting times (around 80 minutes) can be followed by either a long
or short eruption (around 2 minutes).
